Bare minimum: order theory

### 2025-09-06 BareMinima/OrderTheory/BareMinimumOrderTheory.tex
OrderTheory: reviewing a bit

\begin{abstract}
This note presents a condensed summary of order theory which can function as a crash course, refresher and/or reference. Bare minima are meant to give a rough overview, by no means complete, of the subject to the intellectually curious, particularly in the context of foundational questions in physics.
\end{abstract}

@@

The mathematical structures studied by order theory appear in different contexts, making the topic relevant to those interested in foundational questions. Core structures in fields such mathematical logic, topology and measure theory, can be characterized and understood in purely order theoretic terms; characterizing number types (i.e. integers, rational and reals) as different order types makes a better link to physics; physical concepts (e.g. part-whole inclusion) are also naturally captured with order theory.

@@

\emph{This work is part of Assumptions of Physics (\url{https://assumptionsofphysics.org}), a project that aims to identify a handful of physical principles from which the basic laws can be rigorously derived.}

@@

Mathematically, the following is the most convenient and common way to define order relationships.

\begin{defn}[Partial orders and poset]
	A \textbf{partial order} \marginleft{Partial order, poset: $(X, \leq)$} $\leq$ is a binary relationship over a set $X$ that is:
	\begin{description}
		\item[Reflexive] $a \leq a$ for all $a \in X$
		\item[Antisymmetric] if $a \leq b$ and $b \leq a$ then $a = b$ for all $a, b \in X$
		\item[Transitive] if $a \leq b$ and $b \leq c$ then $a \leq c$ for all $a, b, c \in X$.
	\end{description}
	A \textbf{partially ordered set}, or \textbf{poset}, is a tuple $(X, \leq)$ consisting on a set and a partial order.
\end{defn}

\begin{remark}
	To give a sense of the broad range of applicability: sets are ordered by set inclusion ($\{0, 1\} \subseteq \{0, 1, 2\}$); statements are ordered by implication/narrowness (\statement{Snoopy is a dog} $\narrower$ \statement{Snoopy is a mammal}); systems are ordered by is-a-part-of (`piston' $\subseteq$ `engine' $\subseteq$ `car').
\end{remark}

@@

\begin{prop}
	A preorder induces a partial order on the set of equivalence classes defined by the preorder. That is, define $\sim$ on $X$ such that $a \sim b$ if $a \lesssim b$ and $b \lesssim a$. This is an equivalence relationship. Now define $\lesssim$ on $X_{/\sim}$ such that $a_{/\sim} \lesssim b_{/\sim}$ if $a \lesssim b$. This is a preorder.
\end{prop}

@@

Simple orders \marginleft{Hasse diagram}  can be depicted using \textbf{Hasse diagrams}. Each element in the poset is represented by a vertex. A line connects upwards an element to its immediate successor, so that lowest elements are at the bottom.

@@

\begin{defn}
	A \textbf{monotonic function} \marginleft{Monotonic function} (also called \textbf{monotone}, \textbf{increasing} or \textbf{order preserving}) is a map between two posets $(X, \leq_X)$ and $(Y, \leq_Y)$ that respects the order. That is, a map $f : X \to Y$ such that $a \leq_X b \implies f(a) \leq_Y f(b)$.
\end{defn}

\begin{defn}
	An \textbf{order embedding} \marginleft{Order embedding} is a monotonic function that respects the order in both directions. That is, a map $f : X \to Y$ such that $a \leq_X b \iff f(a) \leq_Y f(b)$.
\end{defn}

\begin{defn}
	An \textbf{order isomorphism} \marginleft{Order isomorphism} is an invertible embedding (i.e. bijective monotonic function whose inverse is monotonic). Two posets are \textbf{order isomorophic} if there exist an order isomoprhism between them.
\end{defn}

@@

@@

\begin{defn}[Join and meet]
	The \textbf{join} \marginleft{Join and meet: $\vee, \wedge$} of two elements $a, b \in X$ or of a set $A$, noted $a \vee b$ and $\bigvee A$, is their supremum if it exists. The \textbf{meet} of two elements $a, b \in X$, noted $a \wedge b$, is their infimum, if it exists.
\end{defn}

@@

\begin{defn}[Lattice properties]
	We define\marginleft{Complete, distributive, complemented} the following additional properties on lattices:
	\begin{description}
		\item[Completeness.] A lattice is \textbf{complete} if all subsets have joins and meet, not just finite ones.
		\item[Distributivity.] A lattice is \textbf{distributive} if it satisfies
		\begin{align*}
			a \vee (b \wedge c) = (a \vee b) \wedge (a \vee c) \\
			a \wedge (b \vee c) = (a \wedge b) \vee (a \wedge c).
		\end{align*}
		A complete lattice is \textbf{infinitely distributive} if the same holds over infinite operations.
		\item[Complemented.] A bounded lattice is \textbf{complemented} if every element has a unique complement.
	\end{description}
\end{defn}

>>>>

\begin{abstract}
A summary overview of order theory, which can be used as a reference, refresher or to be aware of what one does not know.
\end{abstract}

@@

Order theory studies partially ordered sets (POSETs): sets equipped with a notion of order. These structures are pervasive in mathematics as play a role in logic, computer science, topology, measure theory, group theory, and beyond. Even number types types (i.e. integers, rational and reals) can be characterized purely on their order structure. A as different order types makes a better link to physics; physical concepts (e.g. part-whole inclusion) are also naturally captured with order theory.

@@

@@

\begin{defn}[Partial orders and poset]
	A \textbf{partial order} \marginleft{Partial order, poset: $(X, \leq)$} $\leq$ is a binary relationship over a set $X$ that is:
	\begin{description}
		\item[Reflexive] $a \leq a$ for all $a \in X$
		\item[Antisymmetric] if $a \leq b$ and $b \leq a$ then $a = b$ for all $a, b \in X$
		\item[Transitive] if $a \leq b$ and $b \leq c$ then $a \leq c$ for all $a, b, c \in X$.
	\end{description}
	A \textbf{partially ordered set}, or \textbf{poset}, is a tuple $(X, \leq)$ consisting of a set and a partial order.
\end{defn}

\begin{remark}
	To give a sense of the broad range of applicability: sets are ordered by set inclusion ($\{0, 1\} \subseteq \{0, 1, 2\}$); statements are ordered by implication/narrowness/specificity (\statement{Snoopy is a dog} $\narrower$ \statement{Snoopy is a mammal}); systems are ordered by is-a-part-of (`piston' $\subseteq$ `engine' $\subseteq$ `car').
\end{remark}

@@

\begin{prop}
	A preorder induces a partial order on the set of equivalence classes defined by the preorder. That is, define $\sim$ on $X$ such that $a \sim b$ if $a \lesssim b$ and $b \lesssim a$. This is an equivalence relationship. Now define $\lesssim$ on $X_{/\sim}$ such that $a_{/\sim} \lesssim b_{/\sim}$ if $a \lesssim b$. This is a partial order.
\end{prop}

@@

Simple orders \marginleft{Hasse diagram}  can be depicted using \textbf{Hasse diagrams}. Each element of the poset is represented by a vertex. A line connects upwards an element to its immediate successor, so that lowest elements are at the bottom.

@@

\begin{defn}
	An \textbf{increasing function} \marginleft{Increasing function} (also called \textbf{order preserving}) is a map between two posets $(X, \leq_X)$ and $(Y, \leq_Y)$ that preserves the order. That is, a map $f : X \to Y$ such that $a \leq_X b \implies f(a) \leq_Y f(b)$. It is \textbf{strictly increasing} if $a \leq_X b \implies f(a) <_Y f(b)$.
\end{defn}

\begin{defn}
	A \textbf{decreasing function} \marginleft{Decreasing function} (also called \textbf{order reversing}) is a map between two posets $(X, \leq_X)$ and $(Y, \leq_Y)$ that reverses the order. That is, a map $f : X \to Y$ such that $a \leq_X b \implies f(b) \leq_Y f(a)$.  It is \textbf{strictly decreasing} if $a \leq_X b \implies f(b) <_Y f(a)$.
\end{defn}

\begin{defn}
	A \textbf{monotonic function} \marginleft{Monotonic function} (also called \textbf{monotone}) is a map that is either increasing or decreasing.
\end{defn}

\begin{defn}
	An \textbf{order embedding} \marginleft{Order embedding} is a monotonic function where the image have the same ordering. That is, a map $f : X \to Y$ such that $a \leq_X b \iff f(a) \leq_Y f(b)$.
\end{defn}

\begin{defn}
	An \textbf{order isomorphism} \marginleft{Order isomorphism} is an invertible embedding (i.e. bijective monotonic function whose inverse is monotonic). Two posets are \textbf{order isomorphic} (or have the same \textbf{order type}) if there exists an order isomorphism between them.
\end{defn}

\subsection{Filters and ideals AFTER}

@@

\begin{remark}
	The upper/lowers bound of $A$ are not necessarily in $A$.
\end{remark}

@@

\begin{defn}[Join and meet]
	The \textbf{join} \marginleft{Join and meet: $\vee, \wedge$} of two elements $a, b \in X$ or of a set $A \subseteq X$, noted $a \vee b$ and $\bigvee A$, is their supremum if it exists. The \textbf{meet} of two elements $a, b \in X$ or of a set $A \subseteq X$, noted $a \wedge b$ and $\bigwedge A$, is their infimum, if it exists.
\end{defn}

@@

\begin{defn}[Lattice properties]
	We define\marginleft{Complete, distributive, complemented} the following additional properties on lattices:
	\begin{description}
		\item[Completeness.] A lattice is \textbf{complete} if all subsets have joins and meet, not just finite ones.
		\item[Distributivity.] A lattice is \textbf{distributive} if it satisfies
		\begin{align*}
			a \vee (b \wedge c) = (a \vee b) \wedge (a \vee c) \\
			a \wedge (b \vee c) = (a \wedge b) \vee (a \wedge c).
		\end{align*}
		A complete lattice is \textbf{infinitely distributive} if the same holds over infinite operations.
		\item[Complemented.] A bounded lattice is \textbf{complemented} if every element has a complement. It is uniquely complement if the complement is unique.
		\item[Ortho-complemented.] A bounded lattice is \textbf{complemented} if every element has a unique complement.
	\end{description}
\end{defn}
